---
title: "Exploring the BRFSS data"
author: "Intzar Singh"
output:
  html_document:
    fig_height: 4
    highlight: pygments
    theme: spacelab
  pdf_document: default
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data



```{r load-data}
load("/Users/Intzar/Downloads/brffs2013.rd")
```



* * *

## Part 1: Data
 The data collected in this survey used a random sampling technique. This allows
 us to see that the results and  valid interpretations of this data can be generalized
 to a larger population. However, we need to note that these results should only
 be generalized to the populations that this sample is representative of. 
 
 Furthermore, we can also rule out any causal factors in this study because there
 is no control group to compare to. In addition to this, observational studies, which
 this survey can be categorized as, can not determine causal relationships. 
 
 However, we may be able to get information in this data that may help us determine
 weather certain factors deserved to be looked at in more detail. These topics
 of further study could greatly aide the CDC in its efforts to predict upcoming
 health trends for communities.
 
 For more Information on the data collection process as provided by the CDC:
 https://www.cdc.gov/brfss/annual_data/2019/pdf/overview-2019-508.pdf
 

* * *

## Part 2: Research questions

**Research quesion 1:**
  The first question that is of interest is to determine whether there is any relationship 
  between the marital status of a respondent, the number of children in their home,
  and the number of days in a month that they felt they were bad mentally.
  
  This question could help us determine if respondents that were married with
  children felt more positive about their mental health when compared to married 
  respondents without children. On the opposite end of this question, we will also 
  be able to see if any relationships may exist between unmarried respondents
  with kids, and their mental health. 

  More formally :
  
  Do respondents with children feel differently about their mental health based
  on their Marital Status? Furthermore, can we determine a trend for married and
  unmarried respondents mental health and the number of children they have?
  
**Research question 2:**
  Next I want to evaluate the relationship between education level and likelihood 
  of owning a home. To do this, we will use income to categorize our respondents
  and isolate the relationship to be between education and home ownership.I believe
  that income would definitely be a factor in home ownership rates, but that question
  is not of interest. 
  
  More formally:
  
  When income brackets are equalized, are college graduates more likely to own
  homes than non college graduates? We can also do some minor analysis along the
  way to see the discrepancy in income between non college graduates and college
  graduates.
  


**Research question 3:**
  The next research question we would like to explore is the amount of times someone
  has gone to the doctor in the past year, based on whether they wear their seat belt
  and have gotten their flu shot. 
  
  What I am interested in testing here, is a parallel in being responsible about
  your health and safety across different fields. Although we may get things that
  are nonsensical or completely unrelated, I am interested in seeing whether someone
  who wears their seat belt and gets their flu shot, is also more likely to visit the
  doctor.
  
  More formally:
  
  Do respondents that wear their seat belt and have gotten a flu shot visit the doctor more
  than respondents that do not fit that criteria?
* * *

## Part 3: Exploratory data analysis

**Research question 1:**
We are going to first filter the data we need out of the bigger set of data, and
clean it up to get rid of any NA values if they exist. 
```{r Filter Our Data for Relevant Info}
q1_data <- brfss2013 %>%
  select(marital, children, menthlth)

q1_data <- q1_data %>%
  filter_at(vars(marital,children,menthlth), all_vars(!is.na(.)))

```



This is all the data that we really want to clean, so we can now move over into 
looking at some statistics about the data, based on wether the participant is 
married or not.

```{r Do a preliminary Wrangling to see Important Info}
q1_data %>%
  group_by(marital) %>%
  summarise(ment_med = median(menthlth), ment_mean= mean(menthlth))
```


From this table we can see that although the data for all members of the marital 
group has a median of 0, their means have quite a bit of variation between them.
We can see that people who are married have a lower average number of days they
felt poorly about their mental health for their group. Furthermore, we can see that
separated individuals post the highest average. 


I now want to look at the distribution of the mental health variable for each of 
these groups and see if we can find something that is of interest to us.

```{r Visualize all the distributions based on the Categorical Variable - Marital}
ggplot(data = q1_data) +
    geom_bar(aes(x = menthlth), colour = "black", fill = "blue") +
    xlab("Days of Poor Mental Health")+
    facet_wrap(~ marital)
 
```


We can see from the above frequency plot that all of the various distributions for each 
type of data are right skewed. Furthermore, the median of all our distributions
seems to be 0, from the graph as well, which can be verified by looking at the
summary table above. 
```{r Create a Box Plot to Visualize some of our Intuitions about the Data Spread and Difference}
ggplot(data = q1_data, aes(x= marital, y= menthlth)) +
  xlab("Marital Status")+
  ylab("Days of Poor Mental health")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.2))+
  geom_boxplot(color = "steelblue")
```


Insights:From the Frequency Plot, and the Box Plot we can see that some variables
might require some further studying. One case of this can be seen in the Separated
and Divorced cases. The wide range of the mental health variable for Separated
couples when compared to those that are divorced, or married, may suggest that there are some
factors that may be leading to more stress. Furthermore, When we compare the 
means of the data, we can further see that there is definitely some discrepancy
between Divorced participants and their feelings on their mental Health, and those
of Married Participants. One Key thing to note however, is that the median for all of these
categorical variables was the same. What this suggests to me is that those participants
who are Separated or Divorced and express poor mental health for at least 1 day
of the month, will express a greater number on average for participants of the
same type who are married. 

We can now turn over into evaluating how the number of children that a participant
may have affects the number of days they express poor mental health. Ideally, we
will get a working version of a visualization that will allow us to establish some
type of relationship between these variables. 


```{r}
q1_data <- q1_data %>%
  mutate(cat_child = ifelse(children<1,"No Childen",ifelse(children<=3, "1-3 Children","More than 3 Children")))
```

```{r Visualize the Relationship between Children and Mental Health with Points}
ggplot(data = q1_data, aes( x= marital, y= menthlth, fill = cat_child))+
  geom_boxplot()+
  xlab("Marital Status")+
  ylab("Days of Poor Mental Health")+
  guides(fill = guide_legend("Children"))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.2))
```


We can see from this graph that when we look at the Marital Status categories that we have 
decided are of importance to us, we can see that Seperated participants that had children
had a higher median for the amount of days that they felt poor mental health. Furthermore,
Separated participants with more than 3 children had not only the highest median, but also the largest 
middle 50th percentile. 

```{r Create a table to see if we can get some Insights from the descriptive statistics }
q1_data %>%
  group_by(cat_child,marital) %>%
  summarise(ment_med = median(menthlth), ment_mean= mean(menthlth), count = n())
```
This table further shows us the discrepancies in how children affect the mean and median
mental health variable for various individuals. 

Lastly I want to see the distribution of the Mental Health Variable grouped by 
how many children a person has. This will allow me to see whether we have a substantial
amount of respondents to make a judgment on if these factors require further 
studying. 

```{r fig.height=5, fig.width=4 }
ggplot(q1_data)+
  geom_bar(aes(x=menthlth), fill = "steelblue")+
  facet_wrap( ~ cat_child, ncol=1)+
  xlab("Days of Poor Mental Health")
```


From the graph above, we can see that the distributions themselves look very similar,
the only major difference between them being the total number of respondents that
fit into these categories. Because these distributions are similar, I believe that
as we explored earlier, the marital status of a person is having an effect on the 
respondents mental health.

As a whole, I do believe that looking at the data in this perspective has helped
us look into how the different Marital Status variables interact with the other 
two variables of our analysis. If I were to take this past the EDA phase, I would
focus in on the Separated, Married, and Divorced Categories and see if there
can be any inferences drawn based on further study. 





**Research question 2:**
We are going to first filter the data we need out of the bigger set of data, and
clean it up to get rid of any NA values if they exist. 


  When income brackets are taken into consideration, are college graduates more likely to own
  homes than non college graduates? We can also do some minor analysis along the
  way to see the discrepancy in income between non college graduates and college
  graduates.

```{r}
q2_data <- brfss2013 %>%
  select(X_incomg,renthom1,X_educag) %>%
  filter_at(vars(X_incomg,renthom1,X_educag),all_vars(!is.na(.)))
 
q2_data %>% 
  group_by(X_incomg) %>%
  summarise( count = n())

q2_data %>% 
  group_by(X_educag) %>%
  summarise( count = n())

q2_data %>% 
  group_by(renthom1) %>%
  summarise( count = n())
```


The above table summaries give us a good idea of the way the variables are laid out across the 
data set. To get a better visualization of this, we will now create a bar graph. 

There are some things to note about this question that make our analysis different:
- Categorical Data was already provided in the form we need, so no code was needed
- All 3 variables being compared are categorical
- We cant do our normal Numeric operations and Descriptions because these data tables
are all categorical. 


Now I want to first plot the distributions of all of these variables. This will be similar 
to the data tables above, but more visually appealing. 

```{r}
ggplot(data = q2_data, aes(x= X_incomg))+
  geom_bar(fill = "steelblue")+
  xlab("Income Bracket")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.2))
  
ggplot(data = q2_data, aes(x= renthom1))+
  geom_bar(fill = "steelblue")+
  xlab("Home Owwnership")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.2))

ggplot(data = q2_data, aes(x= X_educag))+
  geom_bar(fill = "steelblue")+
  xlab("Education Level")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.2))
  
```


The above bar graphs give us a better visual understanding of the dstribution of 
our data, but we are still missing some comparative visualizations that can help us with our question.

Since we are dealing with only categorical variables, I will start by showing a 
contingency table. We will then evaluate the distrubution based on these contigencies with a layered
bar graph. 
```{r fig.height=5,fig.width=5}
q2_data %>%
  group_by(X_educag, X_incomg) %>%
  summarise(count = n())
```


The above data is rather messy, the key question we are trying to see, is based on 
education level, is the proportion of total respondents income higher with more education. We can
try a bar graph next.

```{r fig.height=5,fig.width=5}
ggplot(data=q2_data)+
  geom_bar(aes(x= X_incomg), fill ="steelblue")+
  facet_wrap(~ X_educag, ncol = 1)+
  xlab("Income Level")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.2))
```


From the above data, it becomes quite clear that as you increase your education level
the more likely you are to fall into a higher income category. 

Finally, lets look at the way homeownership is related to these two variables, and then 
lets determine if there is a way to see all three of these variables together. 

```{r fig.height=5,fig.width=5}
ggplot(data=q2_data)+
  geom_bar(aes(x= renthom1), fill ="steelblue")+
  facet_wrap(~ X_incomg, ncol =1)+
  xlab("Income Level")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.2))
```
```{r fig.height=5,fig.width=5}
ggplot(data=q2_data)+
  geom_bar(aes(x= renthom1), fill ="steelblue")+
  facet_wrap(~ X_educag, ncol = 1)+
  xlab("Income Level")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.2))
```


In data table above graphs would look like the following, click through to see the various 
differences
```{r fig.height=5,fig.width=5}
q2_data %>%
  group_by(X_educag,renthom1) %>%
  summarise(count = n())
q2_data %>%
  group_by(X_incomg,renthom1) %>%
  summarise(count = n())

```
```{r fig.height=5,fig.width=8}
ggplot(data=q2_data)+
  geom_bar(aes(x= renthom1, fill = X_incomg))+
  facet_wrap(~ X_educag ,ncol = 1)+
  xlab("Income Level")+
  guides(fill = guide_legend("Income Level"))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.2))
```


Insights: The above data shows that there is clearly a relationship between Education Level,
Income, and Home Ownership. Although the relationship between Education Level, and 
Income may be expected, I was more surprised about the Education Level and Home
Ownership relationship. From our last graph, we can see that Home Ownership Proportions
are higher for those that have completed college than those who have not. This leads
me to believe there may be a third variable, such as loan acceptance rates that may
be confounding this relationship.

This would be a great topic of study to possible gather more data in and try to
generalize this relationship. 


**Research question 3:**
Restating the Question from Above:

 Do respondents that wear their seat belt and have gotten a flu shot visit the doctor more
  than respondents that do not fit that criteria?

First lets discuss the variables that we will be using for this question. The seat belt use
is a ordinal categorical variable, flu shot variable tells us if the respondent has gotten
a flu shot in the last 12 months, and the doctor visits variable tells us how many visits
a person has made to the doctor over the past 12 months. 

Lets extract and clean our data. The good thing for us here is that in our entire data set,
as we have seen above, all we have had to do is remove NA values. We haven't had the need to
fill in missing values or anything else of the sort. However, this time, we want to get rid 
of some extra variable values that we would not like to consider.
```{r}
q3_data <- brfss2013 %>%
  select(seatbelt,drvisits,flushot6) %>%
  filter_at(vars(seatbelt,drvisits,flushot6),all_vars(!is.na(.)))
q3_data <- q3_data %>%
  filter(seatbelt != "Never drive or ride in a car")
```



I want to start by looking at the distributions of all of the variables in question. 
```{r}
ggplot(data = q3_data)+
  geom_bar(aes(x= seatbelt), fill = "steelblue")+
  xlab("Seatbelt worn while Driving")
ggplot(data = q3_data)+
  geom_bar(aes(x= flushot6), fill = "coral")+
  xlab("Recieved Flushot")
ggplot(data = q3_data)+
  geom_bar(aes(x= drvisits), fill = "seagreen")+
  xlab("# of Doctor Visits")
  
```


We can see from the above that, that overwhelmingly people say that they Always wear their seatbelts,
More people have not gotten their flu shot than those who have, and the fistribution for doctor visits 
is right skewed with a peak about 3 visits.

Lets dive more into the relationships between these variables.
```{r}
ggplot(data = q3_data)+
  geom_bar(aes(x= seatbelt), fill = "coral2")+
  facet_wrap(~ flushot6, ncol = 1)+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.2))+
  xlab("Seatbelt worn while Driving")
```


These distributions are similar, so it seems that having gotten your flushot does not relate to the
frequency of wearing your seat belt.
```{r}
ggplot(data = q3_data)+
  geom_bar(aes(x= drvisits), fill = "coral2")+
  facet_wrap(~ flushot6, ncol = 1)+
  xlab("# of Doctor Visits")
  
```


Again, there doesn't seem to be a difference between the data set distribution based on the flu shot 
variable alone. But there does seem to be a shift towards the right in terms of its peak value for
the group that have gotten their flu shot. We can look at a table to have a clearer view. 
```{r}
q3_data%>%
  group_by(flushot6) %>%
  summarise( meanDR= mean(drvisits), medDR = median(drvisits))
```


Here we can see that the Mean and Median are both higher for the Doctor Visits variable, for the group
that has gotten their flu shot. 

We can now try and see if these three variables combined have a relationship that may be of interest to us.
We will be generating a box plot 
```{r}
ggplot(data = q3_data, aes(x= seatbelt, y= drvisits, fill = flushot6)) + 
  geom_boxplot()+
  xlab("Seatbelt worn while Driving")+
  ylab("Doctor Visits")+
  guides(fill = guide_legend("Flushot Recieved"))
```

Insights:

From all of the data visualizations and analysis above, I think one of the more 
interesting questions to study would be if health consciousness can be predicted 
by using variables like Flu Shot. If that is the case we might be able to draw a relationship
between many more interesting things such as health conditions and likelihood of recieving
preventative medicine. 